Expert-In-The-Loop Causal Discovery: Integrating Expert Knowledge

### 2024-08-02 paper/main.tex
Updates

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
\title{Human-In-The-Loop Causal Discovery}
\author{
    %Authors
    % All authors must be in the same font size and format.
    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
    AAAI Style Contributions by Pater Patel Schneider,
    Sunil Issar,\\
    J. Scott Penberthy,
    George Ferguson,
    Hans Guesgen,
    Francisco Cruz\equalcontrib,
    Marc Pujol-Gonzalez\equalcontrib
}
\affiliations{
    %Afiliations
    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
    % If you have multiple authors and multiple affiliations
    % use superscripts in text and roman font to identify them.
    % For example,

@@

Understanding the cause-and-effect relationships between variables is a
fundamental goal in many scientific questions as it reveals the underlying
mechanisms of observed phenomena and informs effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to uncover the underlying causal
structure among random variables from a given observational dataset. The
problem of causal discovery has been explored in both the Directed Acyclic
Graphs (DAGs) and Structural Equation Models (SEMs) literatures, each adopting
different approaches. In the DAG literature, the focus has primarily been on
developing automated algorithms that can learn the causal structure from
observational datasets. In contrast, the SEM literature has focused more
towards developing tools that assist researchers in manually constructing or
modifying models.

% To perform any causal effect identification or estimation using Directed
% Acyclic Graphs (DAGs) or Structural Equation Models (SEMs), researchers need to
% first construct a causal structure commonly known as DAG or path diagram. This
% process of constructing the casual structure from data is known as \emph{Causal
% Discovery}. In the field of DAGs, plenty of algorithms have been developed to
% automatically construct DAGs from data. These algorithm takes different
% approaches to causal discovery such as Constraint-based algorithms such as PC
% and Fast Causal Infenrece (FCI) where the algorithm tries to match the
% conditional independences in the data to the ones impled by the model,
% score-based methods such as Hill-Climb Search, Greedy Equivalence Search (GES)
% where the algorithm tries to find a model that has the best score given some
% scoring metric, and continuous optimization where the problem is formulated as
% a constrained optimization problem such as No Tears.

As a result of these efforts, numerous approaches to automated algorithms for
causal discovery have been proposed in the DAG literature, including
constraint-based method like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}. However, the adaption of these algorithms in applied
fields remain limited, and researchers typically prefer to construct DAGs
manually based on their domain expertise \citep{Tennant2020, Petersen2021}.
Even when using automated algorithms, we often have to rely on manual post hoc
modifications to the model learned by the algorithm because they:

@@

Due to these challenges, whether constructing the model manually or using an
algorithm, we must rely on domain expertise to construct or modify the
model. Hence, it is crucial to have tools to guide this modification process.
One of the ways to guide this modification is to use Conditional Independence
(CI) based model testing \cite{Ankan2021}. As DAGs imply CI statements that can
be tested in the given data using statistical CI tests, we can verify whether
the implied CIs are consistent in the data or not. If some of implied CIs are
not valid in the given dataset, we can modify the model accordingly.
Figure~\ref{fig:ci_testing} shows an example of model testing using CIs.
However, this CI testing approach has a few drawbacks. Firstly, as CIs are
implied only by missing edges in the model, we can not test whether an
incorrect edge is already present in the model. Secondly, the number of implied
CIs can be quite large. For example, in the moderately sized alarm network with 
$ 37 $ node, there are $287$ implied CIs. If we rely only on the p-values of the
CI tests, it can be difficult to figure which test to focus on.

% As it is evident that researchers also draw DAGs by hand and even if they use
% an algorithm, the need to make some post-hoc adjustments to the model, in this
% paper we present a tool to assist researchers in making these adjustments
% somewhat inspired by modification indices. In the case of DAGs, global testing
% methodology does exist such as Shipley test, that can combine local structure
% tests into a single p-value test. However, this does not give us a way to make
% modifications to our model. A possible way to do this by checking which of
% these local tests are violated. However, this approach can quickly become
% diffult as the size of the network increases. For example, the popular alarm
% network implies $x$ condition independence statements.

>>>>

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
\title{Expert-In-The-Loop Causal Discovery: Integrating Expert Knowledge}
\author{
    %Authors
    % All authors must be in the same font size and format.
    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
    AAAI Style Contributions by Pater Patel Schneider,
    Sunil Issar,\\
    J. Scott Penberthy,
    George Ferguson,
    Hans Guesgen,
    Francisco Cruz\equalcontrib,
    Marc Pujol-Gonzalez\equalcontrib
}
\affiliations{
    %Afiliations
    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
    % If you have multiple authors and multiple affiliations
    % use superscripts in text and roman font to identify them.
    % For example,

@@

Understanding the cause-and-effect relationships between variables is a
fundamental goal in many scientific questions as it reveals the underlying
mechanisms of observed phenomena and informs effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to uncover these underlying
cause-effect relationships among random variables from observational data. The
problem of causal discovery has been explored in both the Directed Acyclic
Graphs (DAGs) and Structural Equation Models (SEMs) literatures, each adopting
a different approach. In the DAG literature, the focus has primarily been on
developing automated algorithms that can learn the causal structure from
datasets. In contrast, the SEM literature has focused more towards developing
tools that assist researchers in manually constructing or modifying models.

As a result of these efforts, numerous algorithms for causal discovery have
been proposed in the DAG literature, such as constraint-based methods like the
PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods like Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. However, the adaption of these algorithms in applied fields
remain limited, and researchers typically prefer to construct DAGs manually
based on their domain expertise \citep{Tennant2020, Petersen2021}. We believe this 
is the case because of problems with these algorithms when applied in practical
settings:

@@

Due to these challenges, even when using automated algorithms, we often have to
rely on manual post hoc testing and modifications to the model learned by the
algorithm. Whether constructing the model manually or using an
algorithm, we must rely on domain expertise to construct or modify the
model. Hence, it is crucial to have tools to guide this modification process.
One of the ways to guide this modification is to use Conditional Independence
(CI) based model testing \cite{Ankan2021}. As DAGs imply CI statements that can
be tested in the given data using statistical CI tests, we can verify whether
the implied CIs are consistent in the data or not. If some of implied CIs are
not valid in the given dataset, we can modify the model accordingly.
Figure~\ref{fig:ci_testing} shows an example of model testing using CIs.
However, this CI testing approach has a few drawbacks. Firstly, as CIs are
implied only by missing edges in the model, we can not test whether an
incorrect edge is already present in the model. Secondly, the number of implied
CIs can be quite large. For example, in the moderately sized alarm network with 
$ 37 $ node, there are $287$ implied CIs. If we rely only on the p-values of the
CI tests, it can be difficult to figure which test to focus on.
